Retrofitting Effect Handlers onto OCaml

### 2020-11-13 retro-concurrency.tex
cps

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. If the compiler were to use CPS, then
reifying continuations in the source language is relatively straight-forward.
But OCaml does not use a CPS IR, and changing the OCaml compiler to utilise a CPS IR
would be an enormous undertaking that would affect the
performance profile of existing OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
actively considering effect handlers as one of the mechanism for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project which adds virtual threads and delimited
continuations to Java. Swift concurrency roadmap~\cite{swift} includes
direct-style asynchronous programming and structured concurrency as milestones.
We believe that our design choices will inform similar choices to be made in the
other industrial-strength languages.

>>>>

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. 
%If the compiler were to use CPS, then
%reifying continuations in the source language is relatively straight-forward.
OCaml does not use a CPS IR, and changing the compiler to utilise a CPS IR
would be an enormous undertaking that would affect the
performance profile of existing OCaml applications due to increased memory allocations as
continuation closures get allocated on the heap~\cite{Farvardin20}. Moreover,
with CPS, stack unwinding is meaningless and hence, we would lose compatibility
with tools that inspect the program stack. Hence, we choose not to use CPS
translation and represent continuations as call stacks.

@@

Our work is also timely. The WebAssembly~\cite{Hass17} community group is
considering effect handlers as one of the mechanisms for supporting
concurrency, asynchronous I/O and generators~\cite{WasmProposal}. Project
Loom~\cite{loom} is an OpenJDK project that adds virtual threads and delimited
continuations to Java. The Swift roadmap~\cite{swift} includes
direct-style asynchronous programming and structured concurrency as milestones.
We believe that our design choices will inform similar choices to be made in the
other industrial-strength languages.
